---
title: "scRNAseq_METAFlux_analysis"
output: html_document
date: "2025-09-09"
---

---
title: "scRNAseq_METAFlux_analysis"
output: html_document
date: "2025-09-16"
---

```{r setup}
# Load required libraries
library(Seurat)
library(dplyr)
library(METAFlux)
library(pheatmap)

```

#### View Data from Tutorial
```{r}
# Load example data from METAFlux package
data("sc_test_example")
sc_test_example
# An object of class Seurat 
# 29634 features across 397 samples within 1 assay --> 29634 genes and 397 cells
# Active assay: RNA (29634 features, 0 variable features)
#  2 layers present: counts, data --> raw counts and normalized data

# Active assay is "RNA" by default
# Access raw counts
counts_matrix <- GetAssayData(sc_test_example, slot = "counts")
counts_matrix[1:5, 1:5]  # first 5 genes x 5 cells

# Access normalized/log-transformed data
data_matrix <- GetAssayData(sc_test_example, slot = "data")
data_matrix[1:5, 1:5]

# Access metadata
head(sc_test_example@meta.data)
```

The code below is exactly as written in the METAFlux tutorial but it is giving me an error when I run it
I think this may likely be due to differences in the Seurat version
I am currently using version 5.3.0
The tutorial was published on 10/25/2022, and likely built against Seurat v4
Below, I will install a Seurat v4 version from late 2022 and see if that resolves the issue --> seems to be working!
```{r}
# Remove current version first
remove.packages("Seurat")

# Install specific older version
install.packages("remotes") # if not installed
library(remotes)
install_version("Seurat", version = "4.3.0", repos = "http://cran.us.r-project.org")

# Verify
packageVersion("Seurat")
```


```{r}
table(sc_test_example$Cell_type)

mean_exp=calculate_avg_exp(myseurat = sc_test_example,myident = 'Cell_type',n_bootstrap=3,seed=1)
```


#### Load Data and Preprocess
```{r}
# Read in count matrix (genes as rows, cells as columns)
counts <- read.csv("D:/Users/Jessica/METAFlux/Richards_NatureCancer_GBM_scRNAseq_counts.csv",
                   row.names = 1)
counts[1:5, 1:5]  # view first 5 genes × 5 cells

# Check the dimensions of the count matrix
dim(counts)
# [1] 17647 44712
# There are 17,647 genes and 44,712 cells in the dataset
```

```{r}
# Read in metadata, merge metadata with refined cell type labels
refined_celltype_labels <- read.csv(
  "D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/refined_celltype_labels.csv",
  row.names = 1)
# head(refined_celltype_labels)

metadata <- read.csv("D:/Users/Jessica/METAFlux/Richards_NatureCancer_GBM_scRNAseq_meta.csv",
                     row.names = 1,
                     stringsAsFactors = FALSE)
# head(metadata)

# Merge with refined cell type labels (both have rownames = cell IDs)
merged_meta <- cbind(metadata, refined_celltype_labels[rownames(metadata), , drop = FALSE])
dim(merged_meta)
# [1] 44712    31
head(merged_meta)
```

The cell numbers match up between counts and metadata but the cell names do not
For the counts data it looks like this: G1003.A_T_AAACCTGAGTATCGAA whereas for the metadata it looks like this: G1003-A_T_AAACCTGAGTATCGAA
```{r}
# Fix cell name discrepancies
# Replace hyphens with dots in merged_meta rownames
rownames(merged_meta) <- gsub("-", ".", rownames(merged_meta))

# Check if cell names now match
matches <- colnames(counts) == rownames(merged_meta)
all(matches) 
# [1] TRUE
```


```{r}
# Create Seurat object with raw counts
sc_obj <- CreateSeuratObject(counts = counts)

# Normalize the data
sc_obj <- NormalizeData(sc_obj)

# Add metadata to Seurat object
sc_obj@meta.data <- merged_meta

# Check the object
sc_obj

```


#### Run METAFlux Analysis
I am getting these warning messages when running the code below
Warning messages:
1: Removing 43808 cells missing data for vars requested
2: Removing 43788 cells missing data for vars requested
3: Removing 43747 cells missing data for vars requested

I discovered that my cell names dont match up. 
For the counts data it looks like this: G1003.A_T_AAACCTGAGTATCGAA whereas for the metadata it looks like this: G1003-A_T_AAACCTGAGTATCGAA
Added a code chunck after importing metadata to match up cell names
This fixed the issue!
```{r}
mean_exp <- calculate_avg_exp(myseurat = sc_obj, myident = 'RefinedCellType', n_bootstrap=100,seed=1)
head(mean_exp)
```

```{r}
# Calculate metabolic reaction scores
scores <- calculate_reaction_score(data = mean_exp)
head(scores,4)
```

```{r}
# Calculate the fractions of celltype/clusters
round(table(merged_meta$RefinedCellType) / nrow(merged_meta), 4)
```

```{r}
flux <- compute_sc_flux(num_cell = 6,fraction = c(0.4595, 0.0614, 0.0820, 0.0236, 0.2209, 0.1526),
  fluxscore = scores,
  medium = human_blood)

# Rename the flux value column for clarity
# flux <- flux %>% rename(flux_value = V1)

# Check dimensions of flux results
dim(flux)
# [1] 80140     3

```

```{r}
# Extract cell types from row names in flux results
flux$cell_type <- ifelse(
  grepl("^celltype", rownames(flux)),                       # if starts with "celltype"
  sub("^(celltype \\d+).*", "\\1", rownames(flux)),         # keep "celltype X"
  sub("^(\\S+).*", "\\1", rownames(flux))                   # else, keep everything before first space
)

# Extract the flux reaction name and store in separate column
flux$reaction <- sapply(strsplit(rownames(flux), " "), tail, 1)

# Check unique cell types
unique(flux$cell_type)

head(flux)
```

Map cell types to more descriptive names
```{r}
flux$cell_type <- recode(flux$cell_type,
  "celltype 1" = "Macrophage",
  "celltype 2" = "Microglia",
  "celltype 3" = "Oligodendrocyte",
  "celltype 4" = "T-cell",
  "celltype 5" = "Tumor",
  "celltype 6" = "Unknown"
)

```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Example: df is your dataframe
# Choose the reaction you want to plot
reaction_to_plot <- "HMR_3905"

# Filter the reaction of interest
df_sub <- flux %>% filter(reaction == reaction_to_plot)

# Convert from wide to long format
df_long <- df_sub %>%
  pivot_longer(
    cols = starts_with("V"),   # V1, V2, V3, etc.
    names_to = "replicate",    # new column storing V1, V2, V3
    values_to = "flux_value"   # flux values
  )

# Make boxplot
ggplot(df_long, aes(x = cell_type, y = flux_value, fill = cell_type)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = paste("Flux values for reaction", reaction_to_plot),
       x = "Cell type", y = "Flux value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



