---
title: "scRNAseq_METAFlux_analysis"
output: html_document
date: "2025-09-09"
---

---
title: "scRNAseq_METAFlux_analysis"
output: html_document
date: "2025-09-16"
---

```{r setup}
# Load required libraries
library(Seurat)
library(dplyr)
library(METAFlux)
library(pheatmap)

```

#### View Data from Tutorial
```{r}
# Load example data from METAFlux package
data("sc_test_example")
sc_test_example
# An object of class Seurat 
# 29634 features across 397 samples within 1 assay --> 29634 genes and 397 cells
# Active assay: RNA (29634 features, 0 variable features)
#  2 layers present: counts, data --> raw counts and normalized data

# Active assay is "RNA" by default
# Access raw counts
counts_matrix <- GetAssayData(sc_test_example, slot = "counts")
counts_matrix[1:5, 1:5]  # first 5 genes x 5 cells

# Access normalized/log-transformed data
data_matrix <- GetAssayData(sc_test_example, slot = "data")
data_matrix[1:5, 1:5]

# Access metadata
head(sc_test_example@meta.data)
```

The code below is exactly as written in the METAFlux tutorial but it is giving me an error when I run it
I think this may likely be due to differences in the Seurat version
I am currently using version 5.3.0
The tutorial was published on 10/25/2022, and likely built against Seurat v4
Below, I will install a Seurat v4 version from late 2022 and see if that resolves the issue --> seems to be working!
```{r}
# Remove current version first
remove.packages("Seurat")

# Install specific older version
install.packages("remotes") # if not installed
library(remotes)
install_version("Seurat", version = "4.3.0", repos = "http://cran.us.r-project.org")

# Verify
packageVersion("Seurat")
```


```{r}
table(sc_test_example$Cell_type)

mean_exp=calculate_avg_exp(myseurat = sc_test_example,myident = 'Cell_type',n_bootstrap=3,seed=1)
```


#### Load Data and Preprocess
```{r}
# Read in count matrix (genes as rows, cells as columns)
counts <- read.csv("D:/Users/Jessica/METAFlux/Richards_NatureCancer_GBM_scRNAseq_counts.csv",
                   row.names = 1)
counts[1:5, 1:5]  # view first 5 genes × 5 cells

# Check the dimensions of the count matrix
dim(counts)
# [1] 17647 44712
# There are 17,647 genes and 44,712 cells in the dataset
```

```{r}
# Read in metadata, merge metadata with refined cell type labels
refined_celltype_labels <- read.csv(
  "D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/refined_celltype_labels.csv",
  row.names = 1)
# head(refined_celltype_labels)

metadata <- read.csv("D:/Users/Jessica/METAFlux/Richards_NatureCancer_GBM_scRNAseq_meta.csv",
                     row.names = 1,
                     stringsAsFactors = FALSE)
# head(metadata)

# Merge with refined cell type labels (both have rownames = cell IDs)
merged_meta <- cbind(metadata, refined_celltype_labels[rownames(metadata), , drop = FALSE])
dim(merged_meta)
# [1] 44712    31
head(merged_meta)
```

The cell numbers match up between counts and metadata but the cell names do not
For the counts data it looks like this: G1003.A_T_AAACCTGAGTATCGAA whereas for the metadata it looks like this: G1003-A_T_AAACCTGAGTATCGAA
```{r}
# Fix cell name discrepancies
# Replace hyphens with dots in merged_meta rownames
rownames(merged_meta) <- gsub("-", ".", rownames(merged_meta))

# Check if cell names now match
matches <- colnames(counts) == rownames(merged_meta)
all(matches) 
# [1] TRUE
```


```{r}
# Create Seurat object with raw counts
sc_obj <- CreateSeuratObject(counts = counts)

# Normalize the data
sc_obj <- NormalizeData(sc_obj)

# Add metadata to Seurat object
sc_obj@meta.data <- merged_meta

# Check the object
sc_obj

# Save the Seurat object for future use
saveRDS(sc_obj, file = "D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/sc_obj_with_metadata.rds")
```

```{r}
# Load the Seurat object if needed
sc_obj <- readRDS("D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/sc_obj_with_metadata.rds")
sc_obj
```

#### Run METAFlux Analysis
I am getting these warning messages when running the code below
Warning messages:
1: Removing 43808 cells missing data for vars requested
2: Removing 43788 cells missing data for vars requested
3: Removing 43747 cells missing data for vars requested

I discovered that my cell names dont match up. 
For the counts data it looks like this: G1003.A_T_AAACCTGAGTATCGAA whereas for the metadata it looks like this: G1003-A_T_AAACCTGAGTATCGAA
Added a code chunck after importing metadata to match up cell names
This fixed the issue!
```{r}
mean_exp <- calculate_avg_exp(myseurat = sc_obj, myident = 'RefinedCellType', n_bootstrap=100,seed=1)
head(mean_exp)
```

```{r}
# Calculate metabolic reaction scores
scores <- calculate_reaction_score(data = mean_exp)
head(scores,4)
```

```{r}
# Calculate the fractions of celltype/clusters
round(table(merged_meta$RefinedCellType) / nrow(merged_meta), 4)
```

```{r}
flux <- compute_sc_flux(num_cell = 6,fraction = c(0.4595, 0.0614, 0.0820, 0.0236, 0.2209, 0.1526),
  fluxscore = scores,
  medium = human_blood)

# Rename the flux value column for clarity
# flux <- flux %>% rename(flux_value = V1)

# Check dimensions of flux results
dim(flux)
# [1] 80140     3

```

```{r}
# Extract cell types from row names in flux results
flux$cell_type <- ifelse(
  grepl("^celltype", rownames(flux)),                       # if starts with "celltype"
  sub("^(celltype \\d+).*", "\\1", rownames(flux)),         # keep "celltype X"
  sub("^(\\S+).*", "\\1", rownames(flux))                   # else, keep everything before first space
)

# Extract the flux reaction name and store in separate column
flux$reaction <- sapply(strsplit(rownames(flux), " "), tail, 1)

# Check unique cell types
unique(flux$cell_type)

head(flux)
```

Map cell types to more descriptive names
```{r}
flux$cell_type <- recode(flux$cell_type,
  "celltype 1" = "Macrophage",
  "celltype 2" = "Microglia",
  "celltype 3" = "Oligodendrocyte",
  "celltype 4" = "T-cell",
  "celltype 5" = "Tumor",
  "celltype 6" = "Unknown"
)

```


```{r}
# Save the flux dataframe to a CSV file
write.csv(flux, "D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/scrnaseq_flux_results.csv", row.names = TRUE)
```

```{r}
# Read the flux dataframe from the CSV file
flux <- read.csv("D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/scrnaseq_flux_results.csv", row.names = 1)
head(flux)
```

Make boxplots for each cell type given a reaction
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Example: df is your dataframe
# Choose the reaction you want to plot
# "HMR_3905"
reaction_to_plot <- "HMR_5512"

# Filter the reaction of interest
df_sub <- flux %>% filter(reaction == reaction_to_plot)

# Convert from wide to long format
df_long <- df_sub %>%
  pivot_longer(
    cols = starts_with("V"),   # V1, V2, V3, etc.
    names_to = "replicate",    # new column storing V1, V2, V3
    values_to = "flux_value"   # flux values
  )

# Make boxplot
ggplot(df_long, aes(x = cell_type, y = flux_value, fill = cell_type)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = paste("Flux values for reaction", reaction_to_plot),
       x = "Cell type", y = "Flux value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


#### Significance Testing
The majority of reactions have very significant p-values, as low as e-122.
This is likely due to high variation between cells and the result of having many cell types which further introduces variability
```{r}
library(dplyr)

df_long_full <- flux %>%
  pivot_longer(
    cols = starts_with("V"),   # V1, V2, V3, etc.
    names_to = "replicate",    # new column storing V1, V2, V3
    values_to = "flux_value"   # flux values
  )

dim(df_long_full)
# [1] 8014000       4

results <- df_long_full %>%
  group_by(reaction) %>%
  summarise(
    p_value = kruskal.test(flux_value ~ cell_type)$p.value
  ) %>%
  mutate(fdr_p = p.adjust(p_value, method = "fdr"))

# View results
print(results)

```

Annotate significant results by merging with reaction annotations from the human GEM
```{r}
data("human_gem")
annotated_results <- results %>%
  left_join(human_gem[, c("ID", "EQUATION", "GENE ASSOCIATION", "SUBSYSTEM")],
            by = c("reaction" = "ID"))
```

Filter and annotate significant results
Looking at the top 439 significant results, they were all exchange/demand reactions with no gene associations
Exchange/demand reactions don't directly depend on gene expression, they help to balance out the flux model
```{r}
fdr_cutoff <- 1e-130
sum(annotated_results$fdr_p < fdr_cutoff, na.rm = TRUE)

# Merge significant results with reaction annotations
sig_results <- annotated_results %>% filter(fdr_p < fdr_cutoff)
head(sig_results)
```

Look at only flux that have a gene association
```{r}
fdr_cutoff <- 1e-100
sig_results_with_genes_association <- annotated_results %>% filter(fdr_p < fdr_cutoff) %>% filter(!is.na(`GENE ASSOCIATION`))
head(sig_results_with_genes_association)
```


While it is interesting to have so much cell-type specificity, I am primarily interested in macrophages and tumour cells.
These are also the two cell types that make up the majority of the cells in the dataset.
In the original dataset, macrophages, tumour cells, and unknown cells make up 46%, 22%, and 15% of the cells, respectively.
The combined proportion of tumour and unknown cells will be much closer to that of macrophages.

In the publication, "unknown" and "tumour" labels together match to "malignant" cells
I will now compare macrophages to tumour cells only (combining unknown and tumour labels)
I expect the p-values to be a lot higher, meaning there will be fewer hits

If I combine unknown and tumour into a single tumor category after calculating flux, I get duplicate values for each cell type flux reaction primarily
Therefore, I will combine the two and then recalculate flux
```{r}
# For the RefinedCellType column in sc_obj@meta.data, convert "Unknown" to "Tumor"
sc_obj@meta.data$RefinedCellType <- ifelse(sc_obj@meta.data$RefinedCellType == "Unknown", "Tumor", sc_obj@meta.data$RefinedCellType)
# Check the updated cell type counts
table(sc_obj@meta.data$RefinedCellType)

# Calculate mean expression again with updated cell types
mean_exp <- calculate_avg_exp(myseurat = sc_obj, myident = 'RefinedCellType', n_bootstrap=100,seed=1)
head(mean_exp)
```

```{r}
# Calculate metabolic reaction scores
scores <- calculate_reaction_score(data = mean_exp)
head(scores,4)
```

```{r}
# Calculate the fractions of celltype/clusters
round(table(sc_obj@meta.data$RefinedCellType) / nrow(sc_obj@meta.data), 4)
```

```{r}
flux <- compute_sc_flux(num_cell = 5,fraction = c(0.4595, 0.0614, 0.0820, 0.0236, 0.3735),
  fluxscore = scores,
  medium = human_blood)

# Rename the flux value column for clarity
# flux <- flux %>% rename(flux_value = V1)

# Check dimensions of flux results
dim(flux)
# [1] 67058   100

```

```{r}
# Extract cell types from row names in flux results
flux$cell_type <- ifelse(
  grepl("^celltype", rownames(flux)),                       # if starts with "celltype"
  sub("^(celltype \\d+).*", "\\1", rownames(flux)),         # keep "celltype X"
  sub("^(\\S+).*", "\\1", rownames(flux))                   # else, keep everything before first space
)

# Extract the flux reaction name and store in separate column
flux$reaction <- sapply(strsplit(rownames(flux), " "), tail, 1)

# Check unique cell types
unique(flux$cell_type)

head(flux)
```

Map cell types to more descriptive names
```{r}
flux$cell_type <- recode(flux$cell_type,
  "celltype 1" = "Macrophage",
  "celltype 2" = "Microglia",
  "celltype 3" = "Oligodendrocyte",
  "celltype 4" = "T-cell",
  "celltype 5" = "Tumor",
)

```

```{r}
# Save the flux dataframe to a CSV file
write.csv(flux, "D:/Users/Jessica/METAFlux/computational_flux_analysis/processed/scrnaseq_flux_results_unknown_tumor_combined.csv", row.names = TRUE)
```



Filter and annotate flux results
```{r}
# Filter flux dataframe for only rows were cell_type is Macrophage or Tumor
flux_filtered <- flux %>% filter(cell_type %in% c("Macrophage", "Tumor"))
head(flux_filtered)
dim(flux_filtered)
# [1] 26824    100

```

```{r}
library(dplyr)

df_long_full <- flux %>%
  pivot_longer(
    cols = starts_with("V"),   # V1, V2, V3, etc.
    names_to = "replicate",    # new column storing V1, V2, V3
    values_to = "flux_value"   # flux values
  )

dim(df_long_full)
# [1] 6705800       4

# Perform significance testing between Macrophage and Tumor cell types
results <- df_long_full %>%
  filter(cell_type %in% c("Macrophage", "Tumor")) %>%
  group_by(reaction) %>%
  summarise(
    p_value = wilcox.test(flux_value ~ cell_type)$p.value
  ) %>%
  mutate(fdr_p = p.adjust(p_value, method = "fdr"))

# View results
head(results)
```

Annotate results
```{r}
data("human_gem")
annotated_results <- results %>%
  left_join(human_gem[, c("ID", "EQUATION", "GENE ASSOCIATION", "SUBSYSTEM")],
            by = c("reaction" = "ID"))
```

Look at only flux that have a gene association
```{r}
fdr_cutoff <- 1e-10
sig_results_with_genes_association <- annotated_results %>% filter(fdr_p < fdr_cutoff) %>% filter(!is.na(`GENE ASSOCIATION`))
head(sig_results_with_genes_association)
dim(sig_results_with_genes_association)
# [1] 2898    6
```

```{r}
# Make boxplot for a specific reaction
reaction_to_plot <- "HMR_4931"
# 
# HMR_5332
# HMR_5330
# HMR_4931

# View the full output
# options(tibble.print_max = Inf, tibble.width = Inf)

# Print row of sig_results_with_genes_association for the reaction of interest
print(sig_results_with_genes_association %>% filter(reaction == reaction_to_plot))

df_long_subset <- df_long_full %>%
  filter(cell_type %in% c("Macrophage", "Tumor")) %>%
  filter(reaction == reaction_to_plot)

# Make boxplot
ggplot(df_long_subset, aes(x = cell_type, y = flux_value, fill = cell_type)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = paste("Flux values for reaction", reaction_to_plot),
       x = "Cell type", y = "Flux value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

Search for glutamate related reactions
```{r}
# For the table sig_results_with_genes_association, filter for reactions with "glutamate" in the EQUATION column
glutamate_reactions <- sig_results_with_genes_association %>%
  filter(grepl("glutamate", EQUATION, ignore.case = TRUE))
head(glutamate_reactions)
dim(glutamate_reactions)

```